Unshuffling Permutations

### 2015-12-02 submission.tex
Suite modifications suggerees par les rapporteurs

This paper is intended to study a natural generalization of $\shuffle$,
denoted by $\SHUFFLE$, as a shuffle of permutations. Roughly speaking,
given three permutations $\pi$, $\sigma_1$, and $\sigma_2$, $\pi$ is
said to be a {\em shuffle} of $\sigma_1$ and $\sigma_2$ if $\pi$ (viewed
as a word) is a shuffle of two words that are order-isomorphic to
$\sigma_1$ and $\sigma_2$. Our intention in this paper is to study this
shuffle product of permutations $\SHUFFLE$ both from a combinatorial and
from a computational point of view by focusing on {\em square}
permutations, that are permutations $\pi$ being in the shuffle of a
permutation $\sigma$ with itself. The shuffle product defined and
considered in this paper is different from other existing and well-known
shuffle products on permutations. For instance, in~\cite{DHT:IJAC:2002},
the authors define the {\em shifted shuffle product}. For this product,
$\pi$ is a shuffle of $\sigma_1$ and $\sigma_2$ if $\pi$ is in the
shuffle, as words, of $\sigma_1$ and the word obtained by incrementing
all the letters of $\sigma_2$ by the size of $\sigma_1$. It is a simple
exercise to prove that, given three permutations $\pi$, $\sigma_1$, and
$\sigma_2$, deciding if $\pi$ is a in the shifted shuffle of $\sigma_1$
and $\sigma_2$ is in~\Pclass.
\smallskip

@@

If $u$ is a word of integers without multiple occurrences of a same
letter, $\STD(u)$ is the {\em standarized} of $u$, that is the unique
permutation of the same size as $u$ such that for all $i, j \in [|u|]$,
$u_i < u_j$ if and only if $\STD(u)(i) < \STD(u)(j)$. In particular, the
image of the map $\STD$ is the set $S$ of all permutations. Two words $u$
and $v$ having the same standarized are {\em order-isomorphic}. If
$\sigma$ is a permutation, there is an {\em occurrence} of (the
{\em pattern}) $\sigma$ in $\pi$ if there is a set $P$ of indexes of
letters of $\pi$ such that $\sigma$ and $\pi_{|P}$ are order-isomorphic.
When $\pi$ does not admit any occurrence of $\sigma$, $\pi$ {\em avoids}
$\sigma$.

@@

The purpose of this section is to define a shuffle product $\SHUFFLE$
on permutations. To define this shuffle product, we shall first define
a coproduct denoted by $\Delta$, enabling to unshuffle permutations.
By duality, $\Delta$ implies the definition of $\SHUFFLE$. The reason
why we need to pass by the definition of $\Delta$ to define $\SHUFFLE$
is justified by the fact that a lot of properties of $\SHUFFLE$ depend
of properties of $\Delta$, and that this strategy allows to write concise
and clear proofs of them. We invite the reader unfamiliar with the
concepts of coproduct and duality to consult~\cite{Joni:Rota:1979}
or~\cite{Grinberg:Reiner:2014}.

Let use denote by $\QQ[S]$ the linear span of all permutations. We
define a linear coproduct $\Delta$ on $\QQ[S]$ in the following way. For
any permutation $\pi$, we set
\begin{equation} \label{equ:unshuffling_coproduct}
    \Delta(\pi) =
    \sum_{P_1 \sqcup P_2 = [|\pi|]}
    \STD\left(\pi_{|P_1}\right) \otimes \STD\left(\pi_{|P_2}\right).
\end{equation}
We call $\Delta$ the {\em unshuffling coproduct of permutations}. For
instance,
\begin{equation}
    \Delta(213) =
    \epsilon \otimes 213 + 2 \cdot 1 \otimes 12 +
    1 \otimes 21 + 2 \cdot 12 \otimes 1 + 21 \otimes 1 
    + 213 \otimes \epsilon,
\end{equation}
\begin{equation}
    \Delta(1234) =
    \epsilon \otimes 1234 + 4 \cdot 1 \otimes 123 + 6 \cdot 12 \otimes 12 +
    4 \cdot 123 \otimes 1 + 1234 \otimes \epsilon,
\end{equation}
\begin{equation}\begin{split} \label{equ:example_unshuffling_coproduct}
    \Delta(1432) & =
    \epsilon \otimes 1432 + 3 \cdot {\bf 1 \otimes 132} + 1 \otimes 321 +
    3 \cdot 12 \otimes 21 \\ & + 3 \cdot 21 \otimes 12 
    + 3 \cdot 132 \otimes 1 + 321 \otimes 1 + 1432 \otimes \epsilon.
\end{split}\end{equation}
Observe that the coefficient of the tensor $1 \otimes 132$ is $3$
in~\eqref{equ:example_unshuffling_coproduct} because there are exactly
three ways to extract from the permutation $1432$ two disjoint subwords
respectively order-isomorphic to the permutations $1$ and $132$.

@@

Intuitively, this product shuffles the values and the positions of the
letters of the permutations. As first and easy properties of $\SHUFFLE$,
one can observe that the empty permutation $\epsilon$ is a unit for
$\SHUFFLE$ and that this product is graded by the sizes of the
permutations ({\em i.e.}, the product of a permutation of size $n$ with
a permutation of size $m$ produces a sum of permutations of size $n + m$).
Notice that there are other shuffle products on
permutations~\cite{DHT:IJAC:2002}: the {\em shifted shuffle} and the
{\em convolution product}. Our shuffle product is different from these
last two.

@@

A permutation $\pi$ is a {\em square} if there is a permutation 
$\sigma$ such that $\pi$ appears in $\sigma \SHUFFLE \sigma$.
In this case, we say that $\sigma$ is a {\em square root} of $\pi$.
Equivalently, $\pi$ is a square with $\sigma$ as square root if and only
if in the expansion of $\Delta(\pi)$, there is a tensor
$\sigma \otimes \sigma$ with a nonzero coefficient. In a more
combinatorial way, this is equivalent to say that there are two sets
$P_1$ and $P_2$ of disjoints indexes of letters of $\pi$ satisfying
$P_1 \sqcup P_2 = [|\pi|]$ such that the subwords $\pi_{|P_1}$ and
$\pi_{|P_2}$ are order-isomorphic. Computer experiments give us the
first numbers of square permutations with respects to their size, which
are, from size $0$ to $10$,
\begin{equation}
    1, 0, 2, 0, 20, 0, 504, 0, 21032, 0, 1293418.
\end{equation}
This sequence (and its subsequence obtained by removing the $0$'s) is for
the time being not listed in~\cite{Slo}. The square permutations of
sizes $0$ to $4$ are
\smallskip

@@

Let $u$ be a binary word of length $n$ with $k$ occurrences of $0$.
We denote by $\BINTOPERM$ (Binary word To Permutation) the map sending 
any such word $u$ to the permutation obtained by replacing from left to 
right each occurrence of $0$ in $u$ by $1$, $2$, \dots, $k$, and from 
right to left each occurrence of $1$ in $u$ by $k + 1$, $k + 2$, \dots, $n$.
For instance,
\begin{equation}
    \BINTOPERM({\bf 1}00{\bf 1}0{\bf 1}{\bf 1}0{\bf 1}000) =
    {\bf C} 12 {\bf B} 3 {\bf A} {\bf 9} 4 {\bf 8} 567.
\end{equation}
Observe that for any nonempty permutation $\pi$ in the image of
$\BINTOPERM$, there is exactly one binary word $u$ such that
$\BINTOPERM(u0) = \BINTOPERM(u1) = \pi$. In support of this observation,
when $\pi$ has an even size, we denote by $\PERMTOBIN(\pi)$ (Permutation
To Binary word) the word $ua$ such that $|ua|_0$ and $|ua|_1$ are both 
even, where $a \in \{0, 1\}$.

@@

Let us make an observation about Wilf-equivalence classes of
permutations restrained on square permutations. The
well-known~\cite{Simion:Schmidt:EJC:1985} fact that there is a single
Wilf-equivalence class of permutations of size $3$ together with
Proposition~\ref{prop:square_stability} imply that $123$ and $321$ are
in the same Wilf-equivalence class of square permutations, and that
$132$, $213$, $231$, and $312$ are in the same Wilf-equivalence class of
square permutations. Computer experiments show us that there are two
Wilf-equivalence classes of square permutations of size $3$. Indeed, the
number of square permutations avoiding $123$ begins by
\begin{equation} \label{equ:sequence_square_123}
    1, 0, 2, 0, 12, 0, 118, 0, 1218, 0, 14272,
\end{equation}
while the number of square permutations avoiding $132$ begins by
\begin{equation} \label{equ:sequence_square_132}
    1, 0, 2, 0, 11, 0, 84, 0, 743, 0, 7108.
\end{equation}

@@

\begin{proof}[of Claim~\ref{claim:square_binary_to_square_permutation}]
    Since $u$ is a square binary word, there is a binary word $v$
    such that $u \in v \shuffle v$. Then, there are two disjoint
    sets $P$ and $Q$ of positions of letters of $u$ such that
    $u_{|P} = v = u_{|Q}$. Now, by definition of $\BINTOPERM$, the
    words $\BINTOPERM(u)_{|P}$ and $\BINTOPERM(u)_{|Q}$ have the
    same standarized $\sigma$. Hence, and by definition of
    the shuffle product of permutations, $\BINTOPERM(u)$ appears in
    $\sigma \SHUFFLE \sigma$, showing that $\BINTOPERM(u)$ is a
    square permutation.
    \qed
\end{proof}
\bigskip

>>>>

This paper is intended to study a natural generalization of $\shuffle$,
denoted by $\SHUFFLE$, as a shuffle of permutations. Roughly speaking,
given three permutations $\pi$, $\sigma_1$, and $\sigma_2$, $\pi$ is
said to be a {\em shuffle} of $\sigma_1$ and $\sigma_2$ if $\pi$ (viewed
as a word) is a shuffle of two words that are order-isomorphic to
$\sigma_1$ and $\sigma_2$. This shuffle product was first introduced by
Vargas~\cite{Vargas:2014} under the name of {\em supershuffle}. Our 
intention in this paper is to study this shuffle product of permutations 
$\SHUFFLE$ both from a combinatorial and from a computational point of 
view by focusing on {\em square} permutations, that are permutations 
$\pi$ being in the shuffle of a permutation $\sigma$ with itself. Many 
other shuffle products on permutations appear in the literature. For 
instance, in~\cite{DHT:IJAC:2002}, the authors define the {\em convolution 
product} and the {\em shifted shuffle product}. For this last product, 
$\pi$ is a shuffle of $\sigma_1$ and $\sigma_2$ if $\pi$ is in the shuffle, 
as words, of $\sigma_1$ and the word obtained by incrementing all the 
letters of $\sigma_2$ by the size of $\sigma_1$. It is a simple exercise 
to prove that, given three permutations $\pi$, $\sigma_1$, and $\sigma_2$,
deciding if $\pi$ is in the shifted shuffle of $\sigma_1$ and $\sigma_2$ 
is in~\Pclass.
\smallskip

@@

If $u$ is a word of integers without multiple occurrences of a same
letter, $\STD(u)$ is the {\em standardized} of $u$, that is the unique
permutation of the same size as $u$ such that for all $i, j \in [|u|]$,
$u_i < u_j$ if and only if $\STD(u)(i) < \STD(u)(j)$. In particular, the
image of the map $\STD$ is the set $S$ of all permutations. Two words $u$
and $v$ having the same standardized are {\em order-isomorphic}. If
$\sigma$ is a permutation, there is an {\em occurrence} of (the
{\em pattern}) $\sigma$ in $\pi$ if there is a set $P$ of indexes of
letters of $\pi$ such that $\sigma$ and $\pi_{|P}$ are order-isomorphic.
When $\pi$ does not admit any occurrence of $\sigma$, $\pi$ {\em avoids}
$\sigma$. The set of permutations of size $n$ avoiding $\sigma$ is 
denoted by $S_n(\sigma)$.

@@

The purpose of this section is to define a shuffle product $\SHUFFLE$
on permutations. Recall that a first definition of this product was 
provided by Vargas~\cite{Vargas:2014}. To present an alternative 
definition of this product adapted to our study, we shall first define
a coproduct denoted by $\Delta$, enabling to unshuffle permutations.
By duality, $\Delta$ implies the definition of $\SHUFFLE$. The reason
why we need to pass by the definition of $\Delta$ to define $\SHUFFLE$
is justified by the fact that a lot of properties of $\SHUFFLE$ depend
of properties of $\Delta$, and that this strategy allows to write concise
and clear proofs of them. We invite the reader unfamiliar with the
concepts of coproduct and duality to consult~\cite{Joni:Rota:1979}
or~\cite{Grinberg:Reiner:2014}.

Let us denote by $\QQ[S]$ the linear span of all permutations. We
define a linear coproduct $\Delta$ on $\QQ[S]$ in the following way. For
any permutation $\pi$, we set
\begin{equation} \label{equ:unshuffling_coproduct}
    \Delta(\pi) =
    \sum_{P_1 \sqcup P_2 = [|\pi|]}
    \STD\left(\pi_{|P_1}\right) \otimes \STD\left(\pi_{|P_2}\right).
\end{equation}
We call $\Delta$ the {\em unshuffling coproduct of permutations}. For
instance,
\begin{equation}
    \Delta(213) =
    \epsilon \otimes 213 + 2 \cdot 1 \otimes 12 +
    1 \otimes 21 + 2 \cdot 12 \otimes 1 + 21 \otimes 1 
    + 213 \otimes \epsilon,
\end{equation}
\begin{equation}
    \Delta(1234) =
    \epsilon \otimes 1234 + 4 \cdot 1 \otimes 123 + 6 \cdot 12 \otimes 12 +
    4 \cdot 123 \otimes 1 + 1234 \otimes \epsilon,
\end{equation}
\begin{equation}\begin{split} \label{equ:example_unshuffling_coproduct}
    \Delta(1432) & =
    \epsilon \otimes 1432 + 3 \cdot {\bf 1 \otimes 132} + 1 \otimes 321 +
    3 \cdot 12 \otimes 21 \\ & + 3 \cdot 21 \otimes 12 
    + 3 \cdot 132 \otimes 1 + 321 \otimes 1 + 1432 \otimes \epsilon.
\end{split}\end{equation}
Observe that the coefficient of the tensor $1 \otimes 132$ is $3$
in~\eqref{equ:example_unshuffling_coproduct} because there are exactly
three ways to extract from the permutation $1432$ two disjoint subwords
respectively order-isomorphic to the permutations $1$ and $132$.

@@

Intuitively, this product shuffles the values and the positions of the
letters of the permutations. One can observe that the empty permutation 
$\epsilon$ is a unit for $\SHUFFLE$ and that this product is graded by 
the sizes of the permutations ({\em i.e.}, the product of a permutation 
of size $n$ with a permutation of size $m$ produces a sum of permutations 
of size $n + m$).

@@

A permutation $\pi$ is a {\em square} if there is a permutation 
$\sigma$ such that $\pi$ appears in $\sigma \SHUFFLE \sigma$.
In this case, we say that $\sigma$ is a {\em square root} of $\pi$.
Equivalently, $\pi$ is a square with $\sigma$ as square root if and only
if in the expansion of $\Delta(\pi)$, there is a tensor
$\sigma \otimes \sigma$ with a nonzero coefficient. In a more
combinatorial way, this is equivalent to saying that there are two sets
$P_1$ and $P_2$ of disjoints indexes of letters of $\pi$ satisfying
$P_1 \sqcup P_2 = [|\pi|]$ such that the subwords $\pi_{|P_1}$ and
$\pi_{|P_2}$ are order-isomorphic. Computer experiments give us the
first numbers of square permutations with respects to their size, which
are, from size $0$ to $10$,
\begin{equation}
    1, 0, 2, 0, 20, 0, 504, 0, 21032, 0, 1293418.
\end{equation}
This sequence (and its subsequence obtained by removing the $0$'s) is for
the time being not listed in~\cite{Slo}. The square permutations of
sizes $0$ to $4$ are
\smallskip

@@

Let $u$ be a binary word of length $n$ with $k$ occurrences of $0$.
We denote by $\BINTOPERM$ (Binary word To Permutation) the map sending 
any such word $u$ to the permutation obtained by replacing from left to 
right each occurrence of $0$ in $u$ by $1$, $2$, \dots, $k$, and from 
right to left each occurrence of $1$ in $u$ by $k + 1$, $k + 2$, \dots, $n$.
For instance,
\begin{equation}
    \BINTOPERM({\bf 1}00{\bf 1}0{\bf 1}{\bf 1}0{\bf 1}000) =
    {\bf C} 12 {\bf B} 3 {\bf A} {\bf 9} 4 {\bf 8} 567,
\end{equation}
where $\mathrm{A}$, $\mathrm{B}$, and $\mathrm{C}$ respectively stand 
for $10$, $11$, and $12$. Observe that for any nonempty permutation 
$\pi$ in the image of $\BINTOPERM$, there is exactly one binary word $u$ 
such that $\BINTOPERM(u0) = \BINTOPERM(u1) = \pi$. In support of this
observation, when $\pi$ has an even size, we denote by $\PERMTOBIN(\pi)$
(Permutation To Binary word) the word $ua$ such that $|ua|_0$ and $|ua|_1$ 
are both even, where $a \in \{0, 1\}$.

@@

Let us make an observation about Wilf-equivalence classes of permutations 
restrained on square permutations. Recall that two permutations $\sigma$ 
and $\nu$ of the same size are {\em Wilf equivalent} if 
$\# S_n(\sigma) = \# S_n(\nu)$ for all $n \geq 0$. The
well-known~\cite{Simion:Schmidt:EJC:1985} fact that there is a single
Wilf-equivalence class of permutations of size $3$ together with
Proposition~\ref{prop:square_stability} imply that $123$ and $321$ are
in the same Wilf-equivalence class of square permutations, and that
$132$, $213$, $231$, and $312$ are in the same Wilf-equivalence class of
square permutations. Computer experiments show us that there are two
Wilf-equivalence classes of square permutations of size $3$. Indeed, the
number of square permutations avoiding $123$ begins by
\begin{equation} \label{equ:sequence_square_123}
    1, 0, 2, 0, 12, 0, 118, 0, 1218, 0, 14272,
\end{equation}
while the number of square permutations avoiding $132$ begins by
\begin{equation} \label{equ:sequence_square_132}
    1, 0, 2, 0, 11, 0, 84, 0, 743, 0, 7108.
\end{equation}

@@

\begin{proof}[of Claim~\ref{claim:square_binary_to_square_permutation}]
    Since $u$ is a square binary word, there is a binary word $v$
    such that $u \in v \shuffle v$. Then, there are two disjoint
    sets $P$ and $Q$ of positions of letters of $u$ such that
    $u_{|P} = v = u_{|Q}$. Now, by definition of $\BINTOPERM$, the
    words $\BINTOPERM(u)_{|P}$ and $\BINTOPERM(u)_{|Q}$ have the
    same standardized $\sigma$. Hence, and by definition of
    the shuffle product of permutations, $\BINTOPERM(u)$ appears in
    $\sigma \SHUFFLE \sigma$, showing that $\BINTOPERM(u)$ is a
    square permutation.
    \qed
\end{proof}
\bigskip
